Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-09-01 papers/paper/ms.tex
update ms

\deleted{Classifying Kepler stars as thin vs.\ thick disk members helps connect
them to previous work on the Galaxy's star formation history.  On
extragalactic scales, the star formation rate in spiral galaxies with
mass similar to the Milky Way peaked $\approx$10\,Gyr ago, and has
since decreased by an order of magnitude. }
%\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.}
\added{Despite the expected differences between pencil-beam surveys
and volume-complete samples, our observed age distribution qualitatively
agrees with a number of previous studies.}
\replaced{In our own
galaxy, previous studies have}{Previous work has} measured star formation histories
through CMD fitting of resolved stellar populations
\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X},
and by modeling the local white dwarf luminosity function
\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} and
\citet{2019ApJ...878L..11I} for instance focused on stars (and white
dwarfs) within a few hundred parsecs, and both found a peak in the
local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
focused on stars in a wider 2\,kpc bubble, and additionally reported
three local maxima in the SFR, which they
associated with pericenter passages of the Sagittarius satellite
galaxy.

@@

We began this work with two questions: how wrong is the assumption of
a uniform age distribution for stars in the galactic thin disk?  And
why are only $\approx$50 sub-gigayear transiting planets known, rather
than the $\approx$500 that would be expected under the assumption of a
uniform star formation history for the $\approx$5{,}000 known planets?

@@

Our main conclusions with respect to our original questions are as
follows.

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-0.5ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet hosts show a
    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  \deleted{The star
    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago.}
\item \added{The youngest stars in the Kepler field (0-0.3\,Gyr)
    are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 2.7-3\,Gyr old. The statistical
    uncertainty on this value is dwarfed by the systematic
    uncertainty in the contamination rate of unresolved
    photometric binaries, which could push it as high as 4.81$\pm$0.26 (Section~\ref{subsec:caveats})}.
     \replaced{This result}{This range of values} \replaced{from
    rotation-based ages broadly agrees}{is broadly consistent} with \replaced{recent reports of a
    declining star-formation rate}{modeling expectations (Section~\ref{subsec:trilegal}), and with age distributions derived from} from CMD fitting and white-dwarf
    chronology\replaced{, though with the}{.  The }advantage of \added{rotation-based ages however is their }\replaced{sensitivity}{accuracy} for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the Kepler field
    \replaced{implies}{suggests} that we should instead expect $\approx$250.
  %
  \item We have \replaced{derived}{reported} rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include the youngest
  systems, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode the $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  All quoted age uncertainties are statistical.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}^{\rm \ast}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{
  	EW$_{\rm Li}^{\rm \ast}$ is the lithium equivalent width {\it after} subtracting a constant 7.5\,m\AA\ to account for the \ion{Fe}{1} 6707.44\,\AA\ blend (see Section~\ref{subsec:lithiumsel}).
  	Two checkmarks ($\checkmark\checkmark$) denote confirmed planets with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are
  expected to be reliable.  One checkmark ($\checkmark$) denotes
  confirmed planets with two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before ``candidate'' planets.  
  Planetary sizes are mostly drawn from (in order of precedence)
  \citet{Petigura_2022}, \citet{Berger_2020b_rpage}, and \citet{Thompson_2018}, and might be unphysical
  for grazing planets.
  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main sequence?
  Bit 10: Spectroscopic $|[{\rm Fe/H}]|$$>$0.3?
  Bit 11: \citetalias{Santos_2021} CP/CB?
  Bit 12: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 11 and bit 7.
  The analogous planet quality bitmask, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest a $Q_{\rm star}$ bitmask with bits 0--10
  not raised, and for a planet to be ``reliable'', we suggest
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}

>>>>

\deleted{Classifying Kepler stars as thin vs.\ thick disk members helps connect
them to previous work on the Galaxy's star formation history.  On
extragalactic scales, the star formation rate in spiral galaxies with
mass similar to the Milky Way peaked $\approx$10\,Gyr ago, and has
since decreased by an order of magnitude. }
%\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.}
\added{Despite the expected differences between pencil-beam surveys
and volume-complete samples, our observed age distribution qualitatively
agrees with a number of previous studies.}
\replaced{In our own
galaxy, previous studies have}{Previous work has} measured star formation histories
through CMD fitting of resolved stellar populations
\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X},
and by modeling the local white dwarf luminosity function
\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} \replaced{and
\citet{2019ApJ...878L..11I} for instance focused on stars (and white
dwarfs)}{and \citet{Mazzi2024} for instance focused on stars} within a few hundred parsecs, and both found a peak in the
local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
focused on stars in a wider 2\,kpc bubble, and additionally reported
three local maxima in the SFR, which they
associated with pericenter passages of the Sagittarius satellite
galaxy.

@@

We began this work with two questions: how wrong is the assumption of
a uniform age distribution for stars in the galactic thin disk?  And
why are only $\approx$50 sub-gigayear transiting planets known, rather
than the $\approx$500 that would be expected under the assumption of a
uniform star formation \replaced{history}{rate} for the $\approx$5{,}000 known planets?

@@

\added{Significant care is 
needed when assessing not only the reliability of rotational ages, but also any implications of measured age distributions.
Despite a slew of stellar quality flags, including on metallicity, evolutionary state,
and binarity, we expect that up to a third of the ``apparently clean'' 
Kepler stars with $P_{\rm rot}$$<$10\,days
could potentially be influenced by unresolved binaries (Figure~\ref{fig:hist_tgyro_corr}).
A separate nuance is in understanding that the observed age distribution does not 
directly map onto a star formation history: the kinematic heating of stars in the
disk in particular produces a major bias in the age distribution of stars, and therefore of planets,
observed by pencil-beam surveys such as Kepler.}

Our \deleted{main }conclusions with respect to our original \added{two }questions\added{ -- what is the age distribution in the thin disk, and where are the missing young transiting exoplanets -- }are as
follows.

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-0.5ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet hosts show a
    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  \deleted{The star
    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago.}
\item \added{The youngest stars in the Kepler field (0-0.3\,Gyr)
    are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 2.7-3\,Gyr old. The statistical
    uncertainty on this value is dwarfed by the systematic
    uncertainty in the contamination rate of unresolved
    photometric binaries, which could push it as high as 4.81$\pm$0.26 (Section~\ref{subsec:caveats})}.
     \replaced{This result}{This range of values} \replaced{from
    rotation-based ages broadly agrees}{is broadly consistent} with \replaced{recent reports of a
    declining star-formation rate}{modeling expectations (Section~\ref{subsec:trilegal}), and with age distributions derived from} CMD fitting and white-dwarf
    chronology\replaced{, though with the}{.  The }advantage of \added{rotation-based ages however is their }\replaced{sensitivity}{accuracy} for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the Kepler field
    \replaced{implies}{suggests} that we \replaced{should}{might} instead expect $\approx$250.
  %
  \item We have \replaced{derived}{reported} rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include the youngest
  systems, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode the $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  All quoted age uncertainties are statistical.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}^{\rm \ast}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{
  	EW$_{\rm Li}^{\rm \ast}$ is the lithium equivalent width {\it after} subtracting a constant 7.5\,m\AA\ to account for the \ion{Fe}{1} 6707.44\,\AA\ blend (see Section~\ref{subsec:lithiumsel}).
  	Two checkmarks ($\checkmark\checkmark$) denote confirmed planets with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are
  expected to be reliable.  One checkmark ($\checkmark$) denotes
  confirmed planets with two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before ``candidate'' planets.  
  Planetary sizes are mostly drawn from (in order of precedence)
  \citet{Petigura_2022}, \citet{Berger_2020b_rpage}, and \citet{Thompson_2018}, and might be unphysical
  for grazing planets.
  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main sequence?
  Bit 10: Spectroscopic $|[{\rm Fe/H}]|$$>$0.3?
  Bit 11: \citetalias{Santos_2021} CP/CB?
  Bit 12: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 11 and bit 7.
  The analogous planet quality bitmask, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate disposition not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest a $Q_{\rm star}$ bitmask with bits 0--10
  not raised, and for a planet to be ``reliable'', we suggest
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}
